# Statistical Inference Course Project

### Part 1: Simulation Exercise

In this project I will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. I will investigate the distribution of averages of 40 exponentials. Note that I will need to do a thousand simulations.

#### 1. Show the sample mean and compare it to the theoretical mean of the distribution.

```{r}
# set seed
set.seed(1234)
# set lambda
lambda <- 0.2
# set number of simulations
n <- 1000
# set number of exponentials
m <- 40
# simulate exponential distribution
sim <- replicate(n, rexp(m, lambda))
# calculate mean of each simulation
means <- apply(sim, 2, mean)
```

```{r}
# calculate mean of means
mean(means)
```

```{r}
# calculate theoretical mean
1 / lambda
```

#### 2. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.

```{r}
# calculate variance of means
var(means)
```

```{r}
# calculate theoretical variance
(1 / lambda^2) / m
```

```{r}
# Comparison 
var(means)-(1 / lambda^2) / m
```

Comparing the two values of the **Variance** we see that the values are **very close**.

#### 3. Show that the distribution is approximately normal.

```{r}
# plot histogram of sample distribution and showing both the mean and theoretical mean
library(ggplot2)
# set the ggplot theme
theme_set(theme_dark())
ggplot(data.frame(means), aes(x = means)) + 
  geom_histogram(aes(y = ..density..), binwidth = 0.1, colour = "black", fill = "#ffffff") +
  geom_density(colour = "red") +
  geom_vline(aes(xintercept = mean(means)), colour = "#04ff00", linetype = "solid", size = 0.5) +
  geom_vline(aes(xintercept = 1 / lambda), colour = "#ffb700", linetype = "solid", size = 0.5) +
  ggtitle("Distribution of Sample Means") +
  xlab("Sample Means") +
  ylab("Density")
```

### Part 2 Basic Inferential Data Analysis
