------------------------------------------------------------------------

# Reproducible Research Project 1

## Loading and preprocessing Data By Downloading the Dataset

### Part 1: Load the Data

```{r, echo=TRUE}
# Download the dataset
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
# Create a data directory if it doesn't exist
if (!file.exists("./data")) {
    dir.create("./data")
}
# check if activity.zip file exists
if (!file.exists("./data/activity.zip")) {
    download.file(fileUrl, destfile = "./data/activity.zip")
}
# check if activity.csv file exists
if (!file.exists("./data/activity.csv")) {
    unzip("./data/activity.zip", exdir = "./data")
}
# Read the data
activity <- read.csv("./data/activity.csv")
```

### Part 2 Process/transform the data

**Task 1: What is mean total number of steps taken per day?**  
1. Calculate the total number of steps taken per day.

2\. Make a histogram of the total number of steps taken each day.

3\. Calculate and report the mean and median of the total number of steps taken per day.

```{r, echo=TRUE}
# Calculate the total number of steps taken per day
totalSteps <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)
# Make a histogram of the total number of steps taken each day
hist(totalSteps$steps, breaks = seq(0, 25000, by = 1000), col = "blue", xlab = "Steps", 
    main = "Total Steps per Day")
# Calculate and report the mean and median of the total number of steps taken per day
mean(totalSteps$steps)
median(totalSteps$steps)
```

**Task 2: What is the average daily activity pattern?**  

1.  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
# and the average number of steps taken, averaged across all days (y-axis)
averageSteps <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)
plot(averageSteps$interval, averageSteps$steps, type = "l", main = "Average Steps per Interval", xlab = "Interval", ylab = "Average Steps", col = "red")
# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
averageSteps[which.max(averageSteps$steps), ]$interval
```

**Task 3: Imputing missing values**  

1.  Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2.  Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3.  Create a new dataset that is equal to the original dataset but with the missing data filled in.
4.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


```{r, echo=TRUE}
library(dplyr)
# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
sum(is.na(activity$steps))
# Devise a strategy for filling in all of the missing values in the dataset.
# In codes below, I use mean value for that 5-minute interval for all days.
# Create a new dataset that is equal to the original dataset but with the missing data filled in.
activityImputed <- activity
activity_missing <- activityImputed[is.na(activityImputed$steps), ]
mapped_steps <- merge(activity_missing, averageSteps,by='interval', all.x=TRUE)
# interate activityImputed dataset and replace NA values with mapped_steps$steps by searching same inertval and date
for (i in 1:nrow(activityImputed)) {
    if (is.na(activityImputed$steps[i])) {
        activityImputed$steps[i] <- mapped_steps$steps.y[which(mapped_steps$interval == activityImputed$interval[i] & mapped_steps$date == activityImputed$date[i])]
    }
}
# Create a histogram of the total number of steps taken each day after missing values were imputed

totalStepsImputed <- aggregate(steps ~ date, data = activityImputed, sum)
hist(totalStepsImputed$steps, breaks = seq(0, 25000, by = 1000), col = "blue", xlab = "Steps", 
    main = "Total Steps per Day After Missing Values Imputed")
```

**Task 4: Are there differences in activity patterns between weekdays and weekends? For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.**  

1.  Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
2.  Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE, fig.height=10, fig.width=10}
# Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
# indicating whether a given date is a weekday or weekend day.
activityImputed$day <- ifelse(as.POSIXlt(as.Date(activityImputed$date))$wday%%6 == 
    0, "weekend", "weekday")
# Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis)
# and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

# Create two panel showing plots for weekday and weekend seperatly
par(mfrow = c(2, 1))
# Plot for weekday
averageStepsWeekday <- aggregate(steps ~ interval, data = activityImputed[activityImputed$day ==
    "weekday", ], mean)
plot(averageStepsWeekday$interval, averageStepsWeekday$steps,
    type = "l", main = "Average Steps per Interval for Weekday",
    xlab = "Interval", ylab = "Average Steps", col = "blue"
)
# Plot for weekend
averageStepsWeekend <- aggregate(steps ~ interval, data = activityImputed[activityImputed$day ==
    "weekend", ], mean)
plot(averageStepsWeekend$interval, averageStepsWeekend$steps,
    type = "l", main = "Average Steps per Interval for Weekend",
    xlab = "Interval", ylab = "Average Steps", col = "red"
)

```
